TacticToe: Learning the Game of Tactics

### 2018-01-31 tactictoe_jv.tex
recording: wip

\lstdefinelanguage{SMLSmall}%
{columns=fullflexible,
  keywords={THEN,THENL,val,open,store_thm,fun,fn},%
  frame=none,
  sensitive=true,
  keywordstyle=\fontfamily{lmss}\scriptsize\selectfont,%
  basicstyle=\fontfamily{pcr}\small\selectfont,%
  stringstyle=\tt,
  morestring=[b]",
  literate={\ }{$\mkern6mu$}1%
   {=}{{\tt\raisebox{-.15mm}{=}}}1%
   {[}{{\tt\raisebox{-.15mm}{[}}}1%
   {]}{{\tt\raisebox{-.15mm}{]}}}1%
   {->}{{$\rightarrow$}}1%
   {==>}{{$\Rightarrow$}}1%
   {wedge}{{$\wedge$}}1%
   {<=}{{$\leq$}}1%
   {>=}{{$\geq$}}1%
   {'a}{{$\alpha$}}1%
   {'b}{{$\beta$}}1%
   {ldots}{$\ldots$}1%
   {!}{$\forall$}1%
   {``}{\hspace{-.5mm}`\hspace{-1mm}`}1%
  }
\lstset{language=SML}
\begin{document}
\maketitle

@@

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{llc}
\toprule
 Technique & Parameters & Solved \\
\midrule
Tactic time out & 0.02s & 155 (56.8\%)\\
                & 0.05s (default) & 155 (56.8\%)\\
                & 0.1s & 153 (56.0\%)\\
\midrule
Orthogonalization & $radius = 20$ (default) & 155 (56.8\%)\\ 
                  & $radius = 10$ & 125 (50\%)\\
                  & $radius = 40$ & 125 (50\%)\\
\midrule
Abstraction       & none (default) & 125 (50\%)\\
                  & $theorems = 8$ & 125 (50\%)\\
                  & $theorems = 16$ & 125 (50\%)\\
                  & $theorems = 32$ & 125 (50\%)\\
\midrule
MCTS Policy & $c_{policy} = 1/2$ (default) & 125 (50\%)\\
            & $c_{policy} = 2/3$ & 125 (50\%)\\
            & $c_{policy} = 1/3$ & 125 (50\%)\\
\midrule
MCTS Evaluation & none (default) & 125 (50\%) \\
           & $c_{exploration} =\sqrt{2}$ & 125 (50\%)\\ 
           & $c_{exploration} = 1$ & 125 (50\%)\\ 
           & $c_{exploration} = 2$ & 125 (50\%)\\ 
\midrule
ATP integration & none (default) & 125 (50\%)\\
                & \metis 0.1s & 125 (50\%)\\
                & \metis 0.2s & 125 (50\%)\\
                & \metis 0.4s & 125 (50\%)\\ 
                & asynchronous \eprover & 125 (50\%)\\
                & \metis 0.1s + asynchronous \eprover & 190\\
\midrule
All & best parameters (without \eprover) & \\ 
\bottomrule
\end{tabular}
\end{table}

@@

We first discuss parsing the proofs and identifying 
individual tactic calls. We next show how tactics calls are recorded together 
with their corresponding goals using a modified proof script.
Finally, the recorded data is organized as feature vectors with tactics as 
labels and characteristics of their associated goals as features. 
These feature vectors constitute the training set for our selection algorithm.

Rather than relying on the underlying programming language, we parse the
original theory file containing proof scripts into an
abstract syntax tree (AST).\todoi{reason: reading the proof as if you were a 
human ==> better for the machine learning algorithm (learn the right thing)}

 The AST is then transformed into a globalized AST
where every top-level value is made independent from
the context. \todoi{some reason why this is needed? easy transfer between 
theories} We 
split the proofs into a series of tactic units and wrap them 
with a recording function.
By replaying the proofs produced by the globalized AST, the wrapper can
record the globalized tactic string with its input goal as training examples in
a database.
The combined effect of these steps on a \holfour proof will be illustrated on 
Example~\ref{ex:running}.

\begin{example}\label{ex:running}(Running example)
\small
\begin{lstlisting}[language=SMLSmall,frame=tb]
open boolLib Tactic Prim_rec Rewrite
ldots
val LIST_INDUCT_TAC = INDUCT_THEN list_INDUCT ASSUME_TAC
ldots
val MAP_APPEND = store_thm("MAP_APPEND",
  ``!(f:'a->'b) l1 l2. MAP f (APPEND l1 l2) = APPEND (MAP f l1) (MAP f l2)``,
  STRIP_TAC THEN LIST_INDUCT_TAC THEN ASM_REWRITE_TAC [MAP, APPEND])
\end{lstlisting}
\end{example}

%\subsection{Minimal \sml Code Trees}
%
%
%Using the standard types $string$ and $list$ and their type constructors,
%we define a minimal code tree by the following AST:
%\begin{align*}
%M &=_{def} string \\
%I &=_{def} string\ \vert\ Module(M,I)\\
%D &=_{def} Open(M)\ \vert\  Val(I,E) \\
%E &=_{def} Code(I\ list)\ \vert\   Let (D,E)
%\end{align*}
%
%In this AST, the constructors $Let$, $Val$ and $Open$ correspond to
%the reserved \sml tokens \texttt{let}, \texttt{val} and \texttt{open} 
%respectively. $Module(M,I)$ refers to the \sml token \texttt{M.I} where 
%\texttt{M} is a module (also called structure in \sml) and \texttt{I} is an 
%identifier. $D$ is a declaration, $E$ an expression. Inside a 
%declaration $val(I,E)$, $E$ is a definition for $I$.
%To construct such an AST for a theory file, we rely on a custom parser. It 
%transform a source file into a 
%minimal code tree taking into account the scope of each constructor.
%An additional reason to refrain from the use of the interpreter in our case, 
%is that
%the \polyml compiler needs to execute the code preceding an expression 
%before parsing it, which is significantly slower than a custom parser.

\subsection{Globalization}
%\subsubsection {Unfolding locally declared values}
To transfer \sml values as code across theories in a conservative manner, we 
need to 
globalize \sml identifiers. To this end, we replace each identifier by some 
\sml code containing all the information to obtain the value of this identifier 
without knowing the context. The result is a standalone \sml code executable in 
any \holfour theories which will be interpreted in the same way in the absence 
of side effects. Two techniques are helpful to produce such code. The first one 
is to prefix global identifiers by the module (and sub-modules) in which they 
were declared. And the second one is to replace local identifiers 
by their definitions. Prefixing and 
unfolding are applied recursively in 
the definitions.

We present the tactical proof recording mechanism in details here because it 
can help 
another researcher to produce a tactical export in other interactive theorem 
provers.
We now enumerate common \sml constructions and explain 
independently how we globalize them. We use a stack to keep track of previous 
identifier declaration and their globalization.

%We first start by mentioning additional 
%details about module and value declarations.

\subsubsection{Declaration}

\paragraph{Reserved tokens}

\paragraph{Module}

\paragraph{Global identifier}

\paragraph{Local identifier}

To globalize a identifier, we include 
the code of its declaration in the globalized expression. 
Given a declaration \texttt{fun f v = e} or \texttt{val f = e}, the globalized 
expression for the identifier \texttt{f} is \texttt{let fun f v = e' in f 
end} respectively \texttt{let val f = e' in f end}
The expression \texttt{e'} is the globalized version of \texttt{e}. That is to 
say, it is the expression \texttt{e} where each of its tokens has been replaced 
its globalized expressions as constructed in Section~\ref{}.

In the case of value, we can simplify the globalized expression further and 
replace it by \texttt{e'}.
In the case of a function, \texttt{f} and its argument \texttt{v} are binders. 
So, we prevent their globalization during the construction of \texttt{e'}.
%This approach is slightly more verbose than using an 
%anonymous function construction. But it is prettier as 
%we regroup declarations at the top of the proof expression before returning it
%to the user (see Section~\ref{sec:proofdisplay}).

\paragraph{Bound variables}
A pattern is an \sml expression that 
contains \sml 
constructors (or exceptions in the case of \texttt{handle}) and binding 
variables. Our algorithm can recognize cases where pattern matching appears and 
globalize the expression appropriately.  Let \texttt{p} be a pattern 
and 
\texttt{e} be abd expression, the following constructions can 
occur: \texttt{val p = e}, \texttt{case \_ of p => e}, \texttt{fun f p = e}, 
\texttt{fn p => e}, \texttt{handle p => e}. A binding identifiers in \texttt{p}
is a token that is not a constructor (or exception). To produce the globalized 
version of each construction, we globalize constructors (or exception) in the 
pattern \texttt{p} and we protect binding variables in \texttt{e} to prevent 
their modification. When a value identifier \texttt{x} is declared in a pattern 
\texttt{p}, the globalized expression \texttt{e'} for \texttt{x} becomes 
\texttt{let val p = e' in x 
end}.

\paragraph{Infixity}                                                            
In \sml, it is possible to declare infixity rules for a string $s$ that 
is not prefixed by a module. The infixity applies only when an identifier 
exactly match $s$.

\subsubsection{Expression}

\paragraph{Reserved tokens}

\paragraph{Modules}
Declaring a module is made through the keyword 
\texttt{open}. We can fetch 
the values declared in a module \texttt{m} by first erasing all elements from 
\texttt{PolyML.NameSpace} and then calling \texttt{open m}. Additionally, it 
gives us the information of whether a value is a constructor or an exception. 
This will prove to be particularly useful to handle pattern matching 
construction correctly. During the opening of a module, nested modules can be 
declared and push to the stack. In this way, we can globalize them by prefixing 
them by their parent modules when they appear in the program.

\paragraph{Global identifier}
Global values and functions are prefixed by the modules and submodules where 
they were declared. This information obtained by looking up the first 
identifier that corresponds to the values at the top of the stack.

\paragraph{Local identifier}
Each local identifier is replaced by its globalized expression.

We do not unfold local and global values of type theorem. Indeed, we do not 
want to 
unfold proofs into other proofs. Instead, we fetch theorems by looking for its 
formula in the global database of 
theorems. If it does not exists, we consider the tactic to be local and not 
transferable to other theories.

\paragraph{Bound variables}
Binding variables are not globalized

\paragraph{Infixity}
 Therefore, an identifier \texttt{x} could have been treated 
as infix in an original expression \texttt{E} but when 
we substitute it locally by its 
globalized version \texttt{e'}, \texttt{e'} is now considered with prefix 
infixity in \texttt{E} which makes \texttt{E} an invalid expression. Therefore, 
because we 
want to avoid global changes to \texttt{E}, we encapsulate \texttt{e'} by two 
newly 
defined 
infix guards to reproduce 
the original behaviour.
In total there are 20 possible infix settings \texttt{x} can have in \sml.
Those infix guards are defined to have the 
same 
precedence level $\texttt{P}\in[0,9]$ and same associativity 
$\texttt{A}\in[l,r]$ (right or left) as \texttt{x}.
Definitions for the infix guards depend on their associativity:

\begin{definition}(Code of infix guards)
\begin{lstlisting}[language=SMLSmall,frame=tb]
fun a infixlX_open f = fn x ==> f (a,x)      fun g infixlX_close b = g b
fun a infixrX_open g = g a      fun f infixrX_close b = fn x ==> f (x,b) 
\end{lstlisting}
\end{definition}

The previously decribed transformation affects each identifier in a theory 
file. However, when creating the recording theory file, only proof scripts are 
replaced by their 
globalized version and the rest of the code is left unchanged. The effect on 
the proof script contained in the running example is shown in 
Example~\ref{ex:global}.

\begin{example}\label{ex:global} (Globalization)
\begin{lstlisting}[language=SMLSmall,frame=tb]
Tactic.STRIP_TAC 
  infixl0_open boolLib.THEN infixl0_close
Prim_rec.INDUCT_THEN (DB.fetch "list" "list_INDUCT") Tactic.ASSUME_TAC
  infixl0_open boolLib.THEN infixl0_close
Rewrite.ASM_REWRITE_TAC [DB.fetch "list" "MAP", DB.fetch "list" "APPEND"]
\end{lstlisting}
\end{example}

\subsection{Tactic wrap}

We now restrict the definition of a proof script to being the third 
argument of a call to 
\texttt{store\_thm}. And we precisely define the 
modification applied to the theory file. A modified theory file is identical to 
the original theory file except every proof scripts \texttt{p} is
replaced by \texttt{Split "p'"} where \texttt{Split} is the splitting function 
to be defined and \texttt{"p'"} is the string representation of the globalized 
version of the proof script 
\texttt{p}. 
.
\begin{example}(Proof script transformation)
\begin{lstlisting}[language=SMLSmall,frame=tb]
original file: val id = store_thm (name, term, p)
modified file: val id = store_thm (name, term, Split "p'")
\end{lstlisting}
\end{example}

@@

\begin{example} (Wrap)
\begin{lstlisting}[language=SMLSmall,frame=tb]
R "Tactic.STRIP_TAC"
  infixl0_open boolLib.THEN infixl0_close
R "Prim_rec.INDUCT_THEN (DB.fetch \"list\" \"INDUCT\") Tactic.ASSUME_TAC"
  infixl0_open boolLib.THEN infixl0_close
R "Rewrite.ASM_REWRITE_TAC 
     [DB.fetch \"list\" \"MAP\", DB.fetch \"list\" \"APPEND\"]"
\end{lstlisting}
\end{example}

During its execution, the recording function \texttt{R} (as coded in 
Definition~\ref{ex:record}) 
saves the tactic 
string and the goal, verifies that the tactic string as the same effect as the 
original tactic and apply it to continue the proof. To perform this task, we 
convert the tactic string back into a tactic via the function 
\texttt{read\_tactic}. This function will be used whenever we want to execute a 
tactic string. So in the next sections, we will not make a difference between a 
tactic and its code and call both of them tactics.

\begin{example}\label{ex:record} (Code of the recording function)
\begin{lstlisting}[language=SMLSmall,frame=tb]
fun R tactic_string goal = 
  (save (tactic_string, goal); tactic_of_sml tactic_string goal)
\end{lstlisting}
\end{example}

\subsection{Data Transfer}
We describe here the format preferred to save and load the 
data recorded by the function \texttt{save}. 
During saving, we add fields to the input pair $(tactic,goal)$ to record 
precomputed information in order to make the prediction faster. The final 
feature vector is the tuple $(tactic,time,goal,goal\_list,features)$.
The time $time$ taken by $tactic$ is not currently used by our 
predictor. The list of goals $goal list$ is the output of $tactic$ on 
$goal$. And the features $features$ of the goal $goal$ are extracted and hashed 
as described in Section~\ref{sec:features}.

\begin{remark}
To simplify the description of our machine learning algorithms in future 
sections, we consider a tactic feature vector to be a pair $(tactic,goal)$ and 
only 
explicitly rely on the 
other fields when necessary.
\end{remark}

We now discuss the type of file we want to store our feature vectors on and
how we plan to write and read \sml values. We start by taking 
inspiration from an existing implementation in \holfour.

To avoid the computational cost of re-proving every theorems at the begining of 
each session, theorem values in \holfour are stored.
They are encoded them in to a string by calling \texttt{Term.write\_raw}.
Conversely, the function \texttt{Term.read\_raw} transforms the string back 
into the original term.
To have theorems binded to sml identifiers, it is necessary to create a 
\sml file where those binders are declared. And for simplicity sake, all theory 
data was fed into this \sml file. However, it is
more efficient to store theory data into a text file and make a custom 
parser which can be run faster than the \polyml compiler.
Thanks to this insight, we have 
modified the way in which \holfour theories are saved to include only \sml 
bindings in a \sml file. All other theory data is now stored separately 
into a text file. After integrating the changes, \holfour theories load three 
times faster in average.

Our \tactictoe data management system do not need to create new \sml
bindings for the user. So we can store all \sml values into a text file which 
can be loaded rapidly.
Storing time, tactic string and features is easy since they are respectively 
floats, strings and integers. And because a goal is composed of terms, we can 
re-use the \holfour theory data reading and writing functions 
for the two other fields.

Finally, we mention some limitations that comes with 
the need to record tactics as a string compared to being able to store a \sml 
value. Access to the name of the tactic enhances machine 
learning ability and user interaction. Yet, because some tactics 
depends on the state of \holfour, for example the state of the global 
simpset, a tactic is re-parsed before each proof attempt to be up-to-date.
As a result, a tactic may have different effect on goals than suggested by the 
database which could compromise the prediction accuracy. In practice, this 
should rarely happen. Indeed, most tactics are stateless and an update on a 
stateful tactic is usually improving it.

@@

When \tactictoe finds a proof of a goal, it returns a search tree 
$\mathfrak{T}$ where the root node containing this goal is solved.
In order to transform this search tree into a proof script, we need to extract 
the tactics that contributed to the proof and combine them using tactic 
combinators.
\todoi{Define tactic combinators?}
By the design of the search, a single tactic combinator, \texttt{THENL}, is 
sufficient. It combines a tactic $t$ with a list of subsequent ones, in such a 
way that after $t$ is called, for each created goal a respective 
tactic from the list is called.
Here is a formal description of the transformation defined by mutual 
recursion on goals and nodes of $\mathfrak{T}$ by the respective function 
$P_goal$ and $P_node$:

\begin{equation*}
P_goal(g) =_{def}
\begin{cases}
T(g) & \text{if A(g,T(g)) = \emptyset},\\
T(g) & \texttt{THENL}\ [P(a_1),\ldots,t_{n-1}] & \text{otherwise.}
\end{cases}
%
%
Proof(a) =_def{def}
\begin{cases}
tac & \text{if $t$ is a root and $g$ the unique goal of the root},\\
tac & \text{if $t$ is a leaf},\\
tac\ \texttt{THENL}\ [P(a_1),\ldots,t_{n-1}] & \text{otherwise.}
\end{cases}
\end{equation*}

where $tac$ is the tactic that produced the node, $c$ is the
only successful child of the root and $c_0, \ldots, c_n$ are the 
children of the node produced by the successful tactic.

The readability of the created proof scripts is improved, by replacing
replacing  \texttt{THENL} by \texttt{THEN} when the list has length 1.
Further post-processing such as
removing unnecessary tactics and theorems has been developed and
improve the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.

\paragraph{Minimizing length and time of the proof} 
First a weak minimization is possible when processing the final proof. If the 
tactic \texttt{A THEN B} appears in the proof and has the same effect as 
\texttt{B} then we can drop \texttt{A} from the proof.
Stronger minimization can be obtained by rerunning the prover with feature 
vectors containing only tactics 
from the proof. To obtain a shorter proof, the exploration coefficient should be
increased and the policy coefficient lowered to encourage exploration in the 
width of the proof tree. Additionally, by giving a weaker evaluation to slow 
tactics, we can aim for faster proofs.

\paragraph{Minimizing tactic arguments}
One of the argument that may contain unnecessary information are lists.
Let $t$ be a tactic applied to a goal $g$ containing a list $l$ as argument. 
And let $t'$ be the tactic $t$ where one element $e$ of $l$ as be removed. If 
$t$ and $t'$ have the same effect on $g$ then $t'$ can replace $t$ in the final 
proof.

This is a generalization of the simplest method used for minimizing a list of 
theorems in ``hammers'' \cite{hammers4qed}. The fact that lists are generally small
make this approach feasible.

\paragraph{Prettification}
Without prettification, the returned tactic is barely readable as it contains 
information to guarantees that each \sml subterm is interpreted in the way in 
any context. Our approach here is to strip this information from the tactic and 
test if the tactic has the same effect as the original one to guarantee that we 
are not changing the proof. For efficiency, 
this is performed on each tactic step separately. We strip module prefixes, 
local declaration. If local declaration can't be 
stripped, we try to group all local declarations under a single "let" binding 
which improves readability and prevents duplication of this declarations.

\todoi{Add an example}

%\paragraph{Step by step exploration.}
%Because a human may have better insight on which tactic to chose next, we also 
%provides an interactive mode, where the user is presented the best $n$ 
%predicted tactics for a goal and to be chosen from. The number $n$ is 
%parameter 
%that can be chosen by the user. Those tactics are curated so that only 
%productive tactics are returned. (define productive)

\section{Case Study}
\todoi{Test tactictoe find new examples on new goals and manual exploration 
mode.}
In this section, we compare \tactictoe proof script with human proof script.

>>>>

\lstdefinelanguage{SMLSmall}%
{columns=fullflexible,
  keywords={THEN,THENL,val,open,store_thm,fun,fn},%
  frame=none,
  sensitive=true,
  keywordstyle=\fontfamily{lmss}\scriptsize\selectfont,%
  basicstyle=\fontfamily{pcr}\small\selectfont,%
  stringstyle=\tt,
  morestring=[b]",
  literate={\ }{$\mkern6mu$}1%
   {=}{{\tt\raisebox{-.15mm}{=}}}1%
   {[}{{\tt\raisebox{-.15mm}{[}}}1%
   {]}{{\tt\raisebox{-.15mm}{]}}}1%
   {->}{{$\rightarrow$}}1%
   {==>}{{$\Rightarrow$}}1%
   {wedge}{{$\wedge$}}1%
   {<=}{{$\leq$}}1%
   {>=}{{$\geq$}}1%
   {'a}{{$\alpha$}}1%
   {'b}{{$\beta$}}1%
   {ldots}{$\ldots$}1%
   {!}{$\forall$}1%
   {fff}{{DB.fetch}}1%
   {``}{\hspace{-.5mm}`\hspace{-1mm}`}1%
  }
\lstset{language=SML}
\begin{document}
\maketitle

@@

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{llc}
\toprule
 Technique & Parameters & Solved \\
\midrule
Tactic time out & 0.02s & 154 (56.8\%)\\
                & 0.05s (default) & 156 (56.8\%)\\
                & 0.1s & 154 (56.0\%)\\
\midrule
Orthogonalization & none & 156 \\
                  & $radius = 10$ & 156 \\
                  & $radius = 20$ (default) & 156 \\ 
                  & $radius = 40$ & 156 \\

\midrule
Abstraction       & none (default)  & 156 (\%)\\ 
                  & $theorems = 8$  & 195 (\%)\\
                  & $theorems = 16$ & 199 (72.9\%)\\
                  & $theorems = 32$ & 195 (\%)\\
\midrule
MCTS Policy & $c_{policy} = 0.4$ & 149 (\%)\\
            & $c_{policy} = 0.5$ (default) & 155 (\%)\\
            & $c_{policy} = 0.6$ & 153 (\%)\\
\midrule
MCTS Evaluation & none (default) & 156 (50\%) \\
				& $c_{exploration} = 0.25$ & 153 153 (50\%)\\ 
				& $c_{exploration} = 0.5$ & 153 155 (50\%)\\ 
				& $c_{exploration} = 1$ & 154 (50\%)\\
				& $c_{exploration} =\sqrt{2}$ & 154 154 (50\%)\\ 
				& $c_{exploration} = 2$ & 152 (50\%)\\ 
\midrule
ATP integration & none (default) & 156 (50\%)\\
                & \metis 0.1s & 125 (50\%)\\
                & \metis 0.2s & 125 (50\%)\\
                & \metis 0.4s & 125 (50\%)\\ 
                & asynchronous \eprover & 125 (50\%)\\
                & \metis 0.1s + asynchronous \eprover & 190\\
\midrule
All & best parameters (without \eprover) & \\ 
\bottomrule
\end{tabular}
\end{table}

@@

%We first discuss parsing proof scripts and identifying 
%individual tactic calls. We next show how to record tactics calls are recorded 
%with their corresponding goals using a modified proof script.
%The recorded data which consists of tactic-goal pairs creates the raw 
%knowledge 
%necessary for our tactic selection algorithm.

Rather than relying on the underlying programming language, we parse the
original theory file containing proof scripts. This enables us to extract the 
code of each tactic. 
Working with the string representation of a tactic is better that working with 
its value:
\begin{itemize}
\item Equality between two functions is easy to do from 
their string representation. The direct consequence is that we can avoid 
predicting the same tactic over and over again. 
\item 
The 
results of 
\tactictoe can be returned in a readable form to the user. In this way, the 
user can learn from the feedback, analyze the proof and possibly improve on it.
An other point is that \tactictoe does not need to be installed to run  
proof scripts generated by \tactictoe. In this way, the further development of 
\tactictoe are not affecting the robustness of \holfour.
\item
It is probably difficult (even impossible) to transfer values 
between theories as they are not run in the same \holfour session. In contrast, 
\sml code can be exported, imported.
\item To produce a tactic value form a code is easy in \sml, which can be 
achieved by using a reference to a tactic and update it with the function 
\texttt{use}.
\end{itemize}

In order to transfer our tactic knowledge between theories, we want to be able 
to re-use the code of a tactic recorded in one theory in an 
other. We also would like to make sure that the code is interpretable and its 
interpretation does not change in the other theory.

Even when developing one theory, the context is continually changing: the open 
modules are different, global and local identifiers are defined and the local 
ones can become out of scope. Therefore, it is unlikely that code extracted 
from proof scripts is interpretable in any part of \holfour.
To solve this issue, we prefix each global identifier by the module where it 
was declared and we recursively unfold local declaration until a local 
identifier can be expressed by through code containing only global identifiers.
We call this process globalization. The result is a standalone \sml code 
executable in any \holfour theories which will be interpreted in the same way 
in the absence 
of side effects.

Therefore, we can guarantee that the behaviour of recorded stateless tactics 
will not change. And because an update on a stateful tactic 
generally increases its strength, our prediction algorithm based on past 
examples will still be reasonably effective.

\subsection{Implementation}
We describe in more details the implementation of the recording algorithm which 
consists of 4 phases: proof script extraction, proof script globalization, 
tactic unit wrapping and creation of tactic-goal pairs.

Because of the large number of \sml constructions, we only describe the effect 
of these steps in a running example that is representative of a typical proof 
script. And we comment on the main steps that transforms the original proof 
script into a modified version that is able to record the code of a tactic and 
the goal(s) it processed during the proof. We omit more involved constructions 
such as the globalization of infix operators.

\begin{example}\label{ex:running}(Running example)
\small
\begin{lstlisting}[language=SMLSmall]
open boolLib Tactic Prim_rec Rewrite
ldots
val LIST_INDUCT_TAC = INDUCT_THEN list_INDUCT ASSUME_TAC
ldots
val MAP_APPEND = store_thm("MAP_APPEND",
  ``!(f:'a->'b) l1 l2. MAP f (APPEND l1 l2) = APPEND (MAP f l1) (MAP f l2)``,
  STRIP_TAC THEN LIST_INDUCT_TAC THEN ASM_REWRITE_TAC [MAP, APPEND])
\end{lstlisting}
\end{example}

\begin{example}\label{ex:r0}(Proof script extraction)
\small
\begin{lstlisting}[language=SMLSmall]
STRIP_TAC THEN LIST_INDUCT_TAC THEN ASM_REWRITE_TAC [MAP, APPEND]
\end{lstlisting}
\end{example}

\begin{example}\label{ex:global} (Globalization)
\begin{lstlisting}[language=SMLSmall]
Tactic.STRIP_TAC THEN
Prim_rec.INDUCT_THEN (DB.fetch "list" "list_INDUCT") Tactic.ASSUME_TAC THEN
Rewrite.ASM_REWRITE_TAC [DB.fetch "list" "MAP", DB.fetch "list" "APPEND"]
\end{lstlisting}
\end{example}

@@

\begin{definition}(Tactic unit)
A tactic unit is an \sml expression of type tactics that does not contain a 
infix operator at its root.
\end{definition}

\begin{example}\label{ex:wrap} (Tactic unit wrapping)
\begin{lstlisting}[language=SMLSmall,frame=tb]
R "Tactic.STRIP_TAC" THEN
R "Prim_rec.INDUCT_THEN (DB.fetch \"list\" \"INDUCT\") Tactic.ASSUME_TAC" THEN
R "Rewrite.ASM_REWRITE_TAC 
  [fff \"list\" \"MAP\", fff \"list\" \"APPEND\"]"
\end{lstlisting}
\end{example}

During its execution, the recording function \texttt{R} (as coded in 
Definition~\ref{ex:record}) 
saves the tactic 
string and the goal, verifies that the tactic string as the same effect as the 
original tactic and apply it to continue the proof. To perform this task, we 
convert the tactic string back into a tactic via the function 
\texttt{tactic\_of\_sml}.

\begin{example}\label{ex:record} (Code of the recording function)
\begin{lstlisting}[language=SMLSmall,frame=tb]
fun R stac goal = (save (stac,goal); tactic_of_sml stac goal)
\end{lstlisting}
\end{example}

The function save(stac,goal) save the pair to disk increasing the number of 
tactic-goal pairs in our database.

%\subsection{Data Transfer}
%We describe here the format preferred to save and load the 
%data recorded by the function \texttt{save}. 
%During saving, we add fields to the input pair $(tactic,goal)$ to record 
%precomputed information in order to make the prediction faster. The final 
%feature vector is the tuple $(tactic,time,goal,goal\_list,features)$.
%The time $time$ taken by $tactic$ is not currently used by our 
%predictor. The list of goals $goal list$ is the output of $tactic$ on 
%$goal$. And the features $features$ of the goal $goal$ are extracted and 
%hashed 
%as described in Section~\ref{sec:features}. 
%
%\begin{remark}
%To simplify the description of our machine learning algorithms in future 
%sections, we consider a tactic feature vector to be a pair $(tactic,goal)$ and 
%only 
%explicitly rely on the 
%other fields when necessary.
%\end{remark}
%
%We now discuss the type of file we want to store our feature vectors on and
%how we plan to write and read \sml values. We start by taking 
%inspiration from an existing implementation in \holfour.
%
%To avoid the computational cost of re-proving every theorems at the begining 
%of 
%each session, theorem values in \holfour are stored.
%They are encoded them in to a string by calling \texttt{Term.write\_raw}.
%Conversely, the function \texttt{Term.read\_raw} transforms the string back 
%into the original term.
%To have theorems binded to sml identifiers, it is necessary to create a 
%\sml file where those binders are declared. And for simplicity sake, all 
%theory 
%data was fed into this \sml file. However, it is
%more efficient to store theory data into a text file and make a custom 
%parser which can be run faster than the \polyml compiler.
%Thanks to this insight, we have 
%modified the way in which \holfour theories are saved to include only \sml 
%bindings in a \sml file. All other theory data is now stored separately 
%into a text file. After integrating the changes, \holfour theories load three 
%times faster in average.
%
%Our \tactictoe data management system do not need to create new \sml
%bindings for the user. So we can store all \sml values into a text file which 
%can be loaded rapidly.
%Storing time, tactic string and features is easy since they are respectively 
%floats, strings and integers. And because a goal is composed of terms, we can 
%re-use the \holfour theory data reading and writing functions 
%for the two other fields.
%
%Finally, we mention some limitations that comes with 
%the need to record tactics as a string compared to being able to store a \sml 
%value. Access to the name of the tactic enhances machine 
%learning ability and user interaction. Yet, because some tactics 
%depends on the state of \holfour, for example the state of the global 
%simpset, a tactic is re-parsed before each proof attempt to be up-to-date.
%As a result, a tactic may have different effect on goals than suggested by the 
%database which could compromise the prediction accuracy. In practice, this 
%should rarely happen. Indeed, most tactics are stateless and an update on a 
%stateful tactic is usually improving it.

@@

When \tactictoe finds a proof of a goal $g_{root}$, it returns a search tree 
$\mathfrak{T}$ where the root node containing $g_{root}$ is solved.
In order to transform this search tree into a proof script, we need to extract 
the tactics that contributed to the proof and combine them using tactic 
combinators.
\todoi{Define tactic combinators?}
By the design of the search, a single tactic combinator, \texttt{THENL}, is 
sufficient. It combines a tactic $t$ with a list of subsequent ones, in such a 
way that after $t$ is called, for each created goal a respective 
tactic from the list is called.

Let $T_{sol}$ be a partial function that from a goal $g$ returns a tactic $t$ 
for which $A(g,t)$ is solved in $\mathfrak{T}$.
The proof extraction mechanism is defined by mutual 
recursion on goals and nodes of $\mathfrak{T}$ by the respective function 
$P_{goal}$ and $P_{node}$:

\begin{align*}
P_{goal}(g) &=_{def} T_{sol}(g)\ \texttt{THENL}\ P_{node}(A(g,T_{sol}(g)))\\
P_{node}(a) &=_{def} [P_{goal}(g_1),\ldots,P_{goal}(g_n)]\ \ \ \text{with}\ 
G(a) = g_1,\ldots,g_n\\
\end{align*}

The extracted proof script of $\mathfrak{T}$ is $P_{goal}(g_{root})$.
We minimally improve it by subtituting \texttt{THENL} by \texttt{THEN} when the 
list of goals is a singleton and removing \texttt{THENL []} during the 
extraction phase.
Further post-processing such as
eliminating unnecessary tactics and theorems has been developed and
improve the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.
\todoi{Not sure if this reference is appropriate anymore}

\paragraph{Minimizing length of the proof} 
A fast and simple minimization is applied when processing the final proof. If 
the tactic \texttt{A THEN B} appears in the proof and has the same effect as 
\texttt{B} then we can replace \texttt{A THEN B} by \texttt{B} in the script.
Optionally, stronger minimization can be obtained 
by rerunning \tactictoe with a low prior 
policy coefficient, no prior evaluation and a database of tactics that contains 
tactics from the discovered proof only.

\paragraph{Minimizing tactic arguments}
Let $t$ be a tactic applied to a goal $g$ containing a list $l$ (of theorems) 
as argument. And let $t'$ be the tactic $t$ where one element $e$ of $l$ as be 
removed. If $t$ and $t'$ have the same effect on $g$ then $t'$ can replace $t$ 
in the final proof. This process is repeated for each element of $l$.
This is a generalization of the simplest method used for minimizing a list of 
theorems in ``hammers'' \cite{hammers4qed}.

\paragraph{Prettification}
Without prettification, the returned tactic is barely readable as it contains 
information to guarantees that each \sml subterm is interpreted in the way in 
any context. Since we return the proof at a point where the \sml interpreter is 
a specific state, we can strip unnecessary information in the form of 
module prefixes. If possible, we group all local declarations in the 
script under a single \texttt{let} binding at the start of the script. And we 
replace extended terms by their quoted version.
All in all, if a prettified tactic $t_p$ has the same effect its original 
$t_o$, we replace $t_p$ by $t_o$ in the proof.\\

The total effect on the readability of a proof script is depicted in 
Example~\ref{ex:pretty}.
\todo{run without prettification}
\begin{example}\label{ex:pretty}\ \\
Extracted proof script:
\begin{lstlisting}[language=SMLSmall]
\end{lstlisting}
Minimized and prettified proof script:
\begin{lstlisting}[language=SMLSmall]
\end{lstlisting}
\end{example}

\section{Case Study}
In this section, we compare \tactictoe proof script with human proof script.
